Analysis

### 2022-05-17 manuscript/manuscript.Rmd
More writeup

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].

@@

CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to $\sim50bp$ target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Ohlsson:2001wu]. A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1. It was subsequently refined to include an additional 9bp motif (M2) [@Schmidt:2012uk]. For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif. Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.

@@

CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Rao:2014aa][@Vietri-Rudan:2015tu], and the enrichment of CTCF and cohesin is observed at TAD boundaries  [@Beagan:2020wz][@Stilianoudakis:2020aa]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops

>>>>

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy]. Point mutation and loss of heterozygosity of CTCF is associated with human cancer, identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].

@@

CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Filippova:1996tv][@Ohlsson:2001wu]. A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1. It was subsequently refined to include an additional 9bp motif (M2) [@Schmidt:2012uk]. For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif. Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding. Studies have reported 20,000-50,000 CTCF binding sites within the human genome [@Kim:2007ux][@Nakahashi:2013uw].

@@

CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops

The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr]. Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains. BORIS became progressively specialized to testis-specific expression throughout amniote evolution. CTCFL is normally expressed only in germ cells but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv]. CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv]. CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].
